\documentclass[11pt]{article}
\newcommand{\LabNumber}{6}
\newcommand{\LabTitle}{Arrays}
\newcommand{\LabTopic}{1D and 2D arrays, traversal, accumulation, passing arrays to functions}
\input{common.tex}

\begin{document}
\labheader

\section*{Objectives}
\begin{itemize}[leftmargin=*,nosep]
  \item Declare, initialize, and traverse 1D and 2D arrays.
  \item Pass arrays to functions and understand that the size is \emph{not} carried with them.
  \item Implement basic array algorithms: search, min/max, reverse, frequency count.
  \item Use indices safely (bounds are 0 to \texttt{size} $-1$).
\end{itemize}

\begin{summary}[Quick Reference]
\textbf{Declaration and initialization.}
\begin{lstlisting}
int  a[5] = {1, 2, 3, 4, 5};
int  b[10] = {0};                /* all zeros */
int  m[3][4] = {{1,2,3,4},{5,6,7,8},{9,10,11,12}};
\end{lstlisting}

\textbf{Bounds.}\quad For \texttt{a[N]} valid indices are \texttt{0..N-1}.
Out-of-bounds access is \emph{undefined behavior} -- it may seem to work and still be wrong.\\
\textbf{Passing to functions.}\quad
\begin{lstlisting}
int sum(int a[], int n);    /* size must be passed separately */
\end{lstlisting}
Inside the function, \texttt{sizeof(a)} is \emph{not} the array size -- it is the size of a pointer.\\
\textbf{Common idiom: max scan.}
\begin{lstlisting}
int max = a[0];
for (int i = 1; i < n; i++) if (a[i] > max) max = a[i];
\end{lstlisting}
\end{summary}

\subsection*{Quick Experiments (Try It)}

\begin{tryit}{Partial initialization zeroes the rest}
Run and observe every element:
\begin{lstlisting}
#include <stdio.h>
int main(void) {
    int a[6] = {10, 20, 30};   /* only 3 initializers for 6 elements */
    for (int i = 0; i < 6; i++)
        printf("a[%d] = %d\n", i, a[i]);
    return 0;
}
\end{lstlisting}
\begin{enumerate}[nosep,leftmargin=*]
  \item What are the values of \texttt{a[3]}, \texttt{a[4]}, \texttt{a[5]}?
  \item Change the initializer to \texttt{= \{0\}}. Does the whole array become zero?
  \item Remove the initializer entirely (\texttt{int a[6];}). Are the values still zero?
        (Run it several times. Does the answer change?) What lesson does this teach?
\end{enumerate}
\end{tryit}

\begin{tryit}{Modifying an array inside a function affects the original}
Run this and compare the output to Lab 5's pass-by-value behaviour:
\begin{lstlisting}
#include <stdio.h>
void fill(int a[], int n) {
    for (int i = 0; i < n; i++) a[i] = i * 10;
}
int main(void) {
    int v[5] = {0};
    printf("before fill: ");
    for (int i = 0; i < 5; i++) printf("%d ", v[i]);
    printf("\n");

    fill(v, 5);
    printf("after  fill: ");
    for (int i = 0; i < 5; i++) printf("%d ", v[i]);
    printf("\n");
    return 0;
}
\end{lstlisting}
\begin{enumerate}[nosep,leftmargin=*]
  \item Did \texttt{fill} change \texttt{v} in \texttt{main}?
  \item In Lab 5, changing a plain \texttt{int} parameter inside a function had no effect
        on the caller. Why is an array different?
  \item (Look ahead) Add \texttt{printf("\%p{\textbackslash}n", (void*)a);} inside \texttt{fill} and
        \texttt{printf("\%p{\textbackslash}n", (void*)v);} in \texttt{main}. Are the addresses the same?
\end{enumerate}
\end{tryit}

\begin{tryit}{Out-of-bounds access -- silent and dangerous}
\textbf{Warning:} this invokes undefined behaviour; save your work first.
\begin{lstlisting}
#include <stdio.h>
int main(void) {
    int a[3] = {1, 2, 3};
    for (int i = 0; i <= 5; i++)   /* intentional over-run */
        printf("a[%d] = %d\n", i, a[i]);
    return 0;
}
\end{lstlisting}
\begin{enumerate}[nosep,leftmargin=*]
  \item Does it compile? Does it crash immediately? What does it print for indices 3, 4, 5?
  \item Run the program a second time. Are those out-of-bounds values the same?
  \item Why is this so dangerous even when it seems to ``work''?
\end{enumerate}
\end{tryit}

\section*{In-Lab Exercises (target: 2 hours)}

\begin{labexercise}{Read, Reverse, Print \hfill {\small \itshape (25 min)}}
Read an integer $N$ ($1 \leq N \leq 100$) and then $N$ integers into an array.
Print the array in reverse order on a single line.
\end{labexercise}

\begin{labexercise}{Min, Max, Average \hfill {\small \itshape (25 min)}}
Read an array of up to 50 doubles and compute the minimum, maximum, and average.
Write three separate functions; do not duplicate the loop.
\end{labexercise}

\begin{labexercise}{Linear Search \hfill {\small \itshape (30 min)}}
Write \texttt{int find(int a[], int n, int key)} that returns the index of the first
occurrence of \texttt{key} in the array, or \texttt{-1} if not present. Test from
\texttt{main} with a key that is present and one that is not.
\end{labexercise}

\begin{labexercise}{Matrix Transpose \hfill {\small \itshape (40 min)}}
Read a $3 \times 4$ matrix of integers and print its $4 \times 3$ transpose. Use
two nested \texttt{for} loops. Format each cell with field width 4 so columns align.
\end{labexercise}

\begin{aicollab}
\textbf{Try these prompts with an AI tool:}
\begin{itemize}[leftmargin=*,nosep]
  \item ``In C, what happens when I access \texttt{a[n]} on an array declared as
        \texttt{int a[n]}? What is undefined behavior?''
  \item ``Why does \texttt{sizeof(a)} inside a function not give the array size when
        \texttt{a} is a parameter?''
  \item ``Walk me through selection sort on \texttt{\{5, 3, 8, 1\}}, showing the
        array state after each pass.''
\end{itemize}

\medskip\textbf{Critical check -- verify, don't just accept:}
Ask the AI to write a function that finds the maximum element in an array. Then call
it with: an array of size 0 (n = 0), an array of size 1, and a normal array.
Run the code and check whether it crashes, returns garbage, or handles these edge cases
correctly. Note whether the AI warned you about them in advance.

\medskip\textbf{Know the limit:}
AI tools routinely generate C array code without bounds checking and will not warn you
that out-of-bounds access is undefined behavior -- meaning it may silently produce wrong
results rather than crashing, which is the most dangerous kind of bug.
\end{aicollab}

\takehomebanner

\begin{trace}[Indexing]
What does this print? (Watch the bounds!)
\begin{lstlisting}
#include <stdio.h>
int main(void) {
    int a[] = {10, 20, 30, 40, 50};
    int s = 0;
    for (int i = 1; i < 4; i++) s += a[i];
    printf("%d %d\n", s, a[0] + a[4]);
    return 0;
}
\end{lstlisting}
\end{trace}

\begin{trace}[2D traversal]
Predict the printed output:
\begin{lstlisting}
#include <stdio.h>
int main(void) {
    int m[2][3] = {{1,2,3},{4,5,6}};
    for (int j = 0; j < 3; j++) {
        for (int i = 0; i < 2; i++) printf("%d ", m[i][j]);
        printf("\n");
    }
    return 0;
}
\end{lstlisting}
\end{trace}

\begin{bug}[Off-by-one + uninitialized accumulator]
List all bugs and write the corrected program:
\begin{lstlisting}
#include <stdio.h>
int main(void) {
    int a[5] = {2, 4, 6, 8, 10};
    int sum;
    for (int i = 0; i <= 5; i++) sum += a[i];
    printf("sum = %d\n", sum);
    return 0;
}
\end{lstlisting}
\end{bug}

\begin{writecode}[Count occurrences]
Write a complete program that reads $N$ integers into an array (\(N \leq 100\)),
then reads a target value and prints how many times the target occurs in the array.
\end{writecode}

\begin{writecode}[Reverse in place]
Write a function \texttt{void reverse(int a[], int n)} that reverses the array
\emph{in place} (no auxiliary array). Then write a \texttt{main} that demonstrates it.
\end{writecode}

\begin{drawex}{2-D array memory layout}
The declaration \texttt{int m[3][4]} stores elements in \emph{row-major} order.\\[4pt]
\textbf{Part A -- Grid view.} Draw three rows of four boxes each. Label every box
with its index notation (\texttt{m[0][0]}, \texttt{m[0][1]}, \ldots, \texttt{m[2][3]}).\\[4pt]
\textbf{Part B -- Flat memory view.} Below the grid, draw a single row of 12
consecutive boxes showing how the same elements are laid out in memory, and
indicate the \emph{byte offset} of each box relative to \texttt{m[0][0]}
(assume \texttt{sizeof(int) = 4}).
\end{drawex}

\end{document}
